---
title: "Simple Income Inequality"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This RMarkdown document describes the steps taken to calculate the income inequality ratios per the methodology in Talberth and Weisdorf.

## Load packages
You have to have all packages installed, then load them as follows. Note that the order is important, as they will mask one another.
```{r message=FALSE, warning=FALSE}
library(DescTools)
library(patchwork)
library(fitdistrplus)
library(spatstat)
library(tidyverse)
```

## Functions
It is good practice to start out the script with the functions you will use
```{r}
DMUI=function(x,m){ #DMUI is a function with inputs x (income) and m (the median OR the sufficiency threshold)
  dmui_out=m*log(x/m)+m #income is adjusted, per TW equation
  dmui_out[which(x<=m)]=x[which(x<=m)] #but we do not adjust any incomes less than the median/sufficiency threshhold
  return(dmui_out) #the function returns the the adjusted income
}  
```

## Read in the income data
The data are derived from HI state tax returns (https://tax.hawaii.gov/stats/a5_1annual/a5_5ind_inc/) and data sent directly by DBEDT to the project team. The data report the number of returns per income bracket per year for years 1960-2018, with 2007-2011, 2019, 2020 missing. Note that the reporting brackets differ by year.
```{r message=FALSE, warning=FALSE}
#Read in data as a tibble
inc_ineq <- read_csv("Inc_ineq_cleaned_ready.csv",col_types = "dddcdddddd") #note Bracket_max is a character because of infinite and the annoying way that read_csv works
```

## Wrangle and look at the income data
We have a bit of clean up work to do on the imported income data. For one, the max bracket always refers to infinity (INF) so we need to assign it a number. We have to make sure that that variable is then recognized as a number (above we indicated it was a character to avoid throwing an error during import). We also add some columns to the dataframe that we will need.
```{r message=FALSE, warning=FALSE}
#Set "infinite" upper bound at 1,000,000 and make sure data types are numeric
inc_ineq$Bracket_max[which(inc_ineq$Bracket_max=="Inf")]=1000000
inc_ineq$Bracket_max=as.numeric(inc_ineq$Bracket_max)

#Calculate bin mean and midpoint
inc_ineq$Bracket_width <- inc_ineq$Bracket_max-inc_ineq$Bracket_min
inc_ineq$Bracket_mid <- (inc_ineq$Bracket_max+inc_ineq$Bracket_min)/2
head(inc_ineq) #Displays the top lines of the dataframe, so you can check these variables got added

#Omit years with all zeros by creating a variable that groups inc_ineq by year, summarizes the # of returns, and filters out any years with no data. A new vector is defined with only the (filtered) years
Nyr <- inc_ineq %>% 
  group_by(Year) %>%
  summarize(S=sum(Num_returns)) %>%
  filter(S>0) 
uY=Nyr$Year

```

## Calculating the ratio between Adjusted and Unadjusted income

### Median-based adjustment

We calculate the weighted median, using bin mean and number of returns. We then adjust all brackets above this median using the function we created above. We only adjust the bracket in which the sample median falls if the bracket mean is greater than the sample median; we do not adjust the bracket if its mean is less than the sample median.
```{r}
#Calculate median from AGI_bin_mean centers and Num_returns (weighted median) for years with data
Y_med <- inc_ineq %>%
  filter(Year %in% uY) %>% 
  group_by(Year) %>% 
  summarise(Median = weighted.median(x = AGI_bin_mean, w = Num_returns))

#Add a column to inc_ineq that is the adjusted mean
#Note we call the function DMUI, we feed it 
# (x) the bin mean income from inc_ineq, and
# (m) the weighted median of the entire distribution that year from Y_med
#The match function finds the median data in Y_med where the year matches inc_ineq
inc_ineq$AGI_bin_mean_DMUI=DMUI(x = inc_ineq$AGI_bin_mean,m=Y_med$Median[match(inc_ineq$Year,Y_med$Year)])

#Generate numerical integrals representing the total AGI for both the adjusted and unadjusted incomes
#We do this by multiplying the (adjusted and unadjusted) bin mean by the number of returns
Ysum <- inc_ineq %>% 
  group_by(Year) %>% 
  summarize(AGI_ann=sum(AGI_USD_all),
            AGI_ann2=sum(AGI_bin_mean*Num_returns),
            AGI_annDMUI=sum(AGI_bin_mean_DMUI*Num_returns))

#generate ratios of unadjusted and adjusted integrals
Ysum$DMUI_Ratio=Ysum$AGI_annDMUI/Ysum$AGI_ann

```

We write everything to output files
```{r}
#Write to output file
output_M<-Ysum %>% select(Year,DMUI_Ratio)
#output_M<-Ysum[,c("Year","DMUI_Ratio")]

write_csv(output_M,"DMUI_Ratio_Median.csv")
```

#### Plots - median-based

```{r message=FALSE, warning=FALSE}
#A quick plot to help see the difference between AGI mean and mid point for each bracket. 
#Note that in lower brackets, the mean and the mid point are similar, but in upper brackets, the midpoint falls below the mean. Note that the two  data groups that are in the highest brackets could reflect an artifact of setting the max at $1million.
ggplot(subset(inc_ineq,Num_returns>0), 
       aes(x=Bracket_mid, y=AGI_bin_mean)) +
         geom_point() +
         geom_abline() +
         labs(title = "AGI mean vs bracket mid-point") +
         scale_x_continuous(name = "mid") +
         scale_y_continuous(name = "mean") +
         theme_bw()

#Plot DMUI ratio by year
ggplot(output_M, 
       aes(Year, DMUI_Ratio)) +
    geom_point() +
    geom_path() +
    labs(title = "Diminishing marginal utility of income ratio", subtitle = "Median method") +
    scale_x_continuous(name="Year",limits=c(1960,2020)) +  
    scale_y_continuous(name = "DMUI ratio", limits=c(0.65, 0.85)) +
    theme_bw()
```


##### Plot all years

This plots the actual and adjusted income distributions for all years.
```{r message=FALSE, warning=FALSE}
ggplot(inc_ineq)+
  geom_point(aes(AGI_bin_mean_DMUI,Num_returns),color="blue")+
  geom_path(aes(AGI_bin_mean_DMUI,Num_returns),color="blue")+
  geom_point(aes(AGI_bin_mean,Num_returns))+
  geom_path(aes(AGI_bin_mean,Num_returns))+
  facet_wrap("Year",scales="free_y")+
  scale_x_log10()+
  geom_vline(data=Y_med,aes(xintercept=Median))
```

### Sufficiency threshold-based adjustment

We start by reading in the data for the sufficiency thresholds, and taking a look at the dataset. We had one sufficiency threshold set by ALICE and a second threshold by DBEDT, which is the one we opted to use. The DBEDT one-adult self-sufficiency income threshold (state: $35,143 in 2018) is from the Self Sufficiency Income Standard Report (https://files.hawaii.gov/dbedt/economic/reports/self-sufficiency/self-sufficiency_2018.pdf). We adjusted the 2018 value for inflation to get previous and future years. Other years for which sufficiency thresholds exist (2009, 2011, 2013, 2014, 2016) were not incorporated.

```{r message=FALSE, warning=FALSE}
suff_th <- read_csv("Suff_thresh_INQ_clean.csv")
#head(suff_th)
```

Now we want to run the DMUI function using the sufficiency threshold, and add the ratio to the output dataframe
```{r}
inc_ineq$AGI_ST_DMUI=DMUI(x = inc_ineq$AGI_bin_mean,m=suff_th$DBEDT_S[match(inc_ineq$Year,suff_th$YEAR)])

Ysum_ST <- inc_ineq %>% 
  group_by(Year) %>% 
  summarize(AGI_ann=sum(AGI_USD_all),
            AGI_ann2=sum(AGI_bin_mean*Num_returns),
            AGI_annSTDMUI=sum(AGI_ST_DMUI*Num_returns))
#view(Ysum_ST)

Ysum_ST$DMUI_Ratio_ST=Ysum_ST$AGI_annSTDMUI/Ysum_ST$AGI_ann
```

We create the output files for the sufficiency threshold and for both methods
```{r message=FALSE, warning=FALSE}
output_ST<-Ysum_ST %>% select(Year,DMUI_Ratio_ST)
#output_ST<-Ysum_ST[,c("Year","DMUI_Ratio_ST")]

write_csv(output_ST,"DMUI_Ratio_ST.csv")

#Create a combined data file
output_all <- left_join(output_M, output_ST)
write_csv(output_all,"DMUI_Ratio_both.csv")
```

#### Plots - sufficiency threshold 
We plot the DMUI ratio based on sufficiency threshold
```{r warning=TRUE}
#plot(Ysum_ST$Year,Ysum_ST$DMUI_Ratio_ST)

ggplot(output_ST, 
  aes(x=Year, y=DMUI_Ratio_ST)) +
  geom_point() +
  geom_path() +
  labs(title = "Diminishing marginal utility of income ratio", subtitle = "Sufficiency method") +
  scale_x_continuous(name = "Year",limits=c(1960,2020)) +
  scale_y_continuous(name = "DMUI Sufficiency threshold", limits=c(0.65, 0.85)) +
  theme_bw()

```

This compares the two ratios
```{r message=FALSE, warning=FALSE}
oa_plot=output_all %>% pivot_longer(cols=c("DMUI_Ratio","DMUI_Ratio_ST"))

ggplot(oa_plot, aes(x = Year, y = value,color=name)) +
  geom_point(aes()) +
  geom_path(aes()) +
  scale_color_discrete(name="Threshold Style",type=c("blue","darkgreen"),labels=c("Median","Sufficiency")) +
  labs(x = "Year",
       y = "Ratio of adjusted income to unadjusted income",
       color = "Legend") +
  theme_bw()

```


## GINI for comparison

To build intuition, we examine income inequality using a traditional indicator, the GINI coefficient. For various reasons, GINI is a poor indicator for Hawaii.
```{r message=FALSE, warning=FALSE}
#calculate the GINI to test
Gini_ii <- inc_ineq %>%
  group_by(Year) %>% 
  summarise(G=Gini(x=Bracket_mid, n=Num_returns,na.rm=T))
Gini_ii  

Gini_iiagi <- inc_ineq %>%
  group_by(Year) %>% 
  summarise(G=Gini(x=AGI_bin_mean, n=Num_returns,na.rm=T))
Gini_iiagi 

#Plot
gii=ggplot(Gini_ii,  
       aes(Year,G)) + 
  geom_point()+geom_path() +
  labs(title = "Gini, Calculated with Bin_Midpoint") +
  scale_x_continuous(name="Year",limits=c(1960,2020))+
  scale_y_continuous(name="Gini coefficient",limits=c(.34,.61))+
  theme_bw()

giia=ggplot(Gini_iiagi,  
       aes(Year,G)) + 
  geom_point()+geom_path() +
  labs(title = "Gini, Calculated with AGI_Bin_Mean") +
  scale_x_continuous(name="Year",limits=c(1960,2020))+
  scale_y_continuous(name="Gini coefficient",limits=c(.34,.61))+
  theme_bw()

gii+giia

```

